Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-09-28 Manuscript/manuscript.tex
Updates based on Pavel's comments until III. A.

The measured order parameter change depends on the response of 
the headgroup on the bound charge, i.e. $m_i$ in Eq.~\ref{OPchangeEQ},
and the ion binding affinity. Thus, the former property has to be
well quantified before using the electrometer concept to
analyze binding affinities. This is done experimentally for a wide range
of systems~\cite{seelig87,beschiasvili91}. To calibrate 
the head group order parameter response also in simulations,
we use experimental data for dihexadecyldimethylammonium (C$_{12}$C$_{16}$$^+$N2C$_1$Br$^-$)
in a POPC bilayer \cite{scherer89}. Dihexadecyldimethylammonium
is a cation surfactant having two acyl chains and bearing a unit
charge in the hydrophilic end. Thus, it is expected to locate
in the bilayer similarly to the phospholipids and the molar ratio then 
gives directly the amount of bound unit charge per lipid $X^\pm$ in these systems \cite{scherer87}.

@@

\subsubsection{Simulations of POPC bilayers with cationic surfactants}
An automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium bromide cationic surfactant,

\vspace{0.5cm}
\chemfig{
(-[:0,1.0,,,draw=none]\chemabove{Br}{\scriptstyle\ominus}) -[:0,3.5,,,draw=none]\chemabove{N}{\scriptstyle\oplus} (-[:135]H_3C)(-[:225]H_3C)(-[:315]{(}CH2{)}_{16}-CH_3)(-[:45]{(}CH2{)}_{12}-CH_3)  
} .
\vspace{0.5cm}

AmberTools program \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions 10\%, 20\%, 30\%, 42\% and 50\% of surfactant in the POPC bilayer.
All systems contained 50 POPC molecules per leaflet, 6340 TIP3P water molecules and
6, 14, 21, 35 or 50 surfactants per leaflet. Parameters for Bromium ions were
not available in the standard Gromacs files for Amber force field, thus the
parameters for Chloride ions were used instead.
The systems were simulated for 200~ns using Lipid14 model for POPC where reasonable lipid neighbor exchange occurs.
First 20~ns were omitted from the analysis.

The same systems were also simulated with ECC-lipid model for POPC using the same setup. 
In these simulations the ECC correction was also applied to the cationic surfactant 
by scaling all charges with the same factor as for ECC-lipids, $f_q=0.8$, 
and by using the atom types with reduced $\sigma$ parameters from ECC-lipids.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=7.5cm]{../Fig/POPCstructure.eps}
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.eps}
  \includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    Top: Chemical structure of POPC with definitions of different order parameters calculated.
    Middle: Order parameters of head group, glycerol backbone and sn-1 and sn-2 tails  from simulations with Lipid14 \cite{dickson14} and ECC-lipids models
    compared with experimental order parameters from \cite{ferreira13}.
    Bottom: X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 \cite{dickson14} and ECC-lipids models. 
    } 
    \todo{X-label misses glycerol region. JOE: glycerol belongs to the headgroup part, but, yes, it wouldn't hurt to write it explicitly.}
\end{figure}

@@

In order to validate the newly developed model, ECC-lipids, 
we compared our simulation results without any ions to
NMR order parameters measurements and x-ray scattering form factors 
(Fig.~\ref{simVSexpNOions}). 
Our tail order parameters, as in the original Lipid14 model~\cite{dickson14},
match the experimental values. This is expected as the ECC model does not modify
the already highly optimized parameters of this region in Lipid14. 
\todo{The discussion about acyl chain and headgroup order parameters needs to be
  finished when the value for C2, the scale are fixed in the figure and
  size of the points are fixed in the figure.
  JOE: I think this is fixed now.}
The headgroup order parameters $\alpha$ and $\beta$ and P-N vector angle are slightly larger in the ECC-lipid model in comparison to Lipid14 (Figs.~\ref{simVSexpNOions} and \ref{OrderParameterCHANGESsurf}). 
With the current data we cannot, however, conclude which is the more realistic
conformation since in both models one headgroup order parameter agrees
with experiments within error bars whereas the other does not.
Slight changes are also observed in glycerol backbone order parameters,
but the accuracy of both models is comparable to the state of art lipid models
available in literature \cite{ollila16}. The agreement between the x-ray scattering
form factors (Fig.~\ref{simVSexpNOions}) and the areas per molecule from simulations and experiments (Table~\ref{tab:apls}) 
confirm that the membrane structural properties are well captured. 
\todo{JOE: I think the following is actually in Fig.~\ref{simVSexpNOions}.}
A structural comparison of ECC-lipids with Lipid14 can be found 
in SI along with results with other water models. 
%The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
%by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B_L14-ECCL17_q80_sig89_surf.eps}
  \caption{\label{OrderParameterCHANGESsurf}
    Headgroup order parameter changes and P-N vector orientation as a function of
    cationic surfactant (dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$)
    in PC bilayer from simulations and experiments \cite{scherer89}.
  }
\end{figure}

>>>>

The measured order parameter change depends on the response of 
the headgroup on the bound charge, i.e. $m_i$ in Eq.~\ref{OPchangeEQ},
and the ion binding affinity. Thus, the former property has to be
well quantified before using the electrometer concept to
analyze binding affinities. This is done experimentally for a wide range
of systems~\cite{seelig87,beschiasvili91}. To calibrate 
the head group order parameter response also in simulations,
we use experimental data for dihexadecyldimethylammonium (2C$_{16}$$^+$N2C$_1$Br$^-$)
in a POPC bilayer \cite{scherer89}. Dihexadecyldimethylammonium
is a cation surfactant having two acyl chains and bearing a unit
charge in the hydrophilic end. Thus, it is expected to locate
in the bilayer similarly to the phospholipids and the molar ratio then 
gives directly the amount of bound unit charge per lipid $X^\pm$ in these systems \cite{scherer87}.

@@

\subsubsection{Simulations of POPC bilayers with cationic surfactants}
An automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium \\[0.5cm]
%\vspace{0.5cm}
\chemfig{
 -[:0,3.5,,,draw=none]\chemabove{N}{\scriptstyle\oplus} (-[:135]H_3C)(-[:225]H_3C)(-[:315]{(}CH2{)}_{16}-CH_3)(-[:45]{(}CH2{)}_{16}-CH_3)
} , 
\vspace{0.5cm} \\
which is one of the cationic surfactants used to experimentally quantify the electrometer concept~\cite{scherer89}.
The AmberTools program \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
the acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions of 10\%, 20\%, 30\%, 42\%, or 50\% of surfactant in the POPC bilayer.
All systems contained 50 POPC molecules per leaflet, 6340 TIP3P water molecules and
6, 14, 21, 35, or 50 surfactants per leaflet. Chloride counter ions were used
in simulations, because parameters for bromide, which was used in the experiment \cite{scherer89},
were not available in the standard Gromacs files for Amber force field.
The systems were simulated for 200~ns using Lipid14 model for POPC where reasonable lipid neighbor exchange occurs.
First 20~ns were omitted from the analysis.

The same systems were also simulated with the ECC-lipid model for POPC using the same setup. 
In these simulations the ECC correction was also applied to the cationic surfactant 
by scaling all charges with the same factor as for ECC-lipids, i.e., $f_q=0.8$, 
and by using the atom types with reduced $\sigma$ parameters from ECC-lipids.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=7.5cm]{../Fig/POPCstructure.eps}
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.eps}
  \includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    Top: Chemical structure of POPC with definitions of different order parameters calculated.
    Middle: Order parameters of head group, glycerol backbone and sn-1 and sn-2 tails  from simulations with Lipid14 \cite{dickson14} and ECC-lipids models
    compared with experimental order parameters from \cite{ferreira13}.
    Bottom: X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 \cite{dickson14} and ECC-lipids models. 
    } 
  \todo{X-label misses glycerol region. JOE: glycerol belongs to the headgroup part, but, yes, it wouldn't hurt to write it explicitly. SAMULI: I would maybe remove the whole text, tell the information in caption and increase the order parameter figure size in z-direction.}
  \todo{Also acyl chain points should correspond the legends and maybe have the same size as HG and glycerol.}
  \todo{Legends would be more clear without multiple points.}
\end{figure}

@@

In order to validate the structural properties of the newly developed ECC-lipids model
we first compared our simulation results without any ions to
NMR order parameters measurements and x-ray scattering form factors 
(Fig.~\ref{simVSexpNOions}). 
As in the original Lipid14 model~\cite{dickson14}, the acyl chain order parameters agree
well with the experimental values. Notably,

and in almost all the other available
lipid models \cite{ollila16}

%This is expected as the ECC model does not modify
%the already highly optimized parameters of this region in Lipid14. 
The headgroup order parameters $\alpha$ and $\beta$ and P-N vector angle are slightly larger
in the ECC-lipid model in comparison to Lipid14 (Figs.~\ref{simVSexpNOions} and \ref{OrderParameterCHANGESsurf}). 
With the current data we cannot, however, conclude which is the more realistic
conformation since in both models one headgroup order parameter agrees
with experiments within error bars whereas the other does not.
Slight changes are also observed in glycerol backbone order parameters,
but the accuracy of both models is comparable to the state of art lipid models
available in literature \cite{ollila16}. The agreement between the x-ray scattering
form factors (Fig.~\ref{simVSexpNOions}) and the areas per molecule from simulations and experiments (Table~\ref{tab:apls}) 
confirm that the membrane structural properties are well captured. 
\todo{JOE: I think the following is actually in Fig.~\ref{simVSexpNOions}.}
A structural comparison of ECC-lipids with Lipid14 can be found 
in SI along with results with other water models. 
%The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
%by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B_L14-ECCL17_q80_sig89_surf.eps}
  \caption{\label{OrderParameterCHANGESsurf}
    Headgroup order parameter changes and P-N vector orientation as a function of
    cationic surfactant (dihexadecyldimethylammonium bromide, 2C$_{16}$$^+$N2C$_1$Br$^-$)
    in PC bilayer from simulations and experiments \cite{scherer89}.
  }
\end{figure}
